---
title: "Analytical Hull-White Implementation Using R"
author: "Mario Santoro"
date: '2018-07-13'
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
  word_document: default
---


#Introduction
ATM's: Affine Term Structure Models

# Hull-White Analytical Model
The Hull–White (1990) general extensions of Vasicek  can be fitted to the initial yield and volatility curve. However, this flexibility has its price: the model cannot be handled analytically in general. We therefore restrict to the following extension of the Vasicek model that was analyzed by Hull and White  in 1994 (analytic HW model):

$$dr(t) = (b(t)+ar(t))dt + \sigma dW(t)$$

$a$ and $\sigma$ are chosen to fit volatility structure whereas $b(t)$ is chosen in order to match the initial yield curve.
The model can be represented also as:
$$ r(t)=x(t)+\alpha(t)$$
where
$$dx(t)=-ax(t)dt+\sigma dW(t)$$
$$x(0)=0$$
$$\alpha(t)=fwd(0,t)+\frac{\sigma^2}{2a^2}\left(1-e^{-at}\right)^2.$$
$fwd(0,t)$ is the *market-implied istanteneous forward rate* at maturity $t>0$.


Analytical HW provide an affine term structure because the bond prices are of the form

$$ P(t,T) = A(t,T)\exp{(B(t,T)r(t))}$$

with 
$$B(t,T)=\frac{1}{a}(e^{-a(T-t)}-1)$$ 
and 
$$A(t,T)=\frac{P(0,T)}{P(0,t)}\exp{\left(-B(t,T)fwd(0,t)-\frac{1}{2}B^2(t,T)\phi(t)\right)}$$
where
$$
\phi(t)=\frac{\sigma^2}{2a}\left(1-\exp(-2at)\right)
$$
We are interested in the *continuously compounded spot rate* $R(t,T)=\frac{-\log(P(t,T))}{T-t}$ and, in the case of analytic HW, we can obtain it explicitly:

$$\begin{aligned}
R(t,T) &= \\
&=\frac{-\log(P(t,T))}{T-t}\\
&= \frac{1}{T-t}\left(-\log(\frac{P(0,T)}{P(0,t)}) +\frac{B(t,T)^2}{2}\phi(t)+B(t,T)fwd(0,t)-B(t,T)r(t) \right)\\
&= \frac{1}{T-t}\left(B(t,T)\left(-r(t)+fwd(0,t)+B(t,T)\frac{\phi(t)}{2}\right)+ 
                  R(0,T)T - R(0,t)t \right)
\end{aligned}
$$
For a more formal derivation you can see [Holmgaard](http://studenttheses.cbs.dk/bitstream/handle/10417/4138/anders_baek_holmgaard.pdf?sequence=1) and [Lund](http://www.jesperlund.com/lecturenotes/ctmod1_2.pdf)

We want to simulate 10 scenarios of $R(t,T)$ every month, starting from next month to 1 years, with maturities every 1 month for 1 years from a specific yield curve.  

```{r,eval=T}
freq <- "monthly"
delta_t <- 1/12
horizon <- 1 # I take horizon = 1 years
nb.sims <- 10 # Scenarios' Number 
mat.max <- 1 #maturities max
```


##Spot Term Structure Estimation

We estimate the spot term structure of interest rates using **termstrc** package.

```{r,eval=TRUE}
suppressPackageStartupMessages(library(termstrc))

maturities0 <- c(3/12,6/12,1:3,
                 seq(5,30, by = 5))
yields <- t(as.matrix(c(-0.210376,-0.137938,-0.027532,              -0.204925,-0.173079,-0.014275,0.545688,
                        0.905779,1.049415,1.087227,
                        1.091134)/100))
yield_curve <- data.frame("Maturities" = maturities0,
                         "Yields" = as.vector(yields))
library(ggplot2)
library(dplyr)
yield_curve %>% ggplot(aes(x = Maturities,
                            y = Yields)) +
  geom_line()
dates0 <- c(as.Date(Sys.Date()) + 13)
zydf <- zeroyields(maturities0,yields,dates0)
zyest <- estim_nss(zydf, method = "asv",
                   tauconstr =  c(0.2, 7, 0.1))
times <- seq(from = 0, to = horizon, by = delta_t)
beta <- zyest$optparam
fwr <- fwr_asv(beta,times)
```
  
## Dimensional Analysis  

Before starting  the estimation and the simulation, we want to do some consideration on measure unities of parameters and functions, in particular regarding the time.  
From $B(t,T)$ and $R(t,T)$ definition:  

- $a$ is $[time]^{-1}$  

- $B(t,T)$ is $[time]$    

- $fwd(0,t)$ is $[time]^{-1}$  

- $phi(t)$ is  $[time]^2$  

- $\sigma^2$ is $[time]^-3$

##Short-term rates simulation  
Now  is time to set the parameters and simulate $r(t)$ using  package ESGtoolkit from [Moudiki](https://cran.r-project.org/web/packages/ESGtoolkit/)

```{r,eval=TRUE}
############# Hull-White short-rates simulaton ####
# Simulation of gaussian shocks with ESGtoolkit

set.seed(4)
eps <- ESGtoolkit::simshocks(n = nb.sims, horizon = horizon, 
                             frequency = "monthly")
# Simulation of the factor x with ESGtoolkit
 a <- .01
 sigma <- a*.2
x <- ESGtoolkit::simdiff(n = nb.sims, horizon = horizon, 
                         frequency = "monthly",  
                         model = "OU", 
                         x0 = 0, theta1 = 0, 
                         theta2 = a, 
                         theta3 = sigma,
                         eps = eps)
# I use forward rates. With the low monthly frequency
# I consider them as being instantaneous forward rates
fwdrates <- ts(replicate(nb.sims, fwr), 
               start = start(x), 
               deltat = deltat(x))
#curves$forwards instead of fwr if you use Rquantlib estimation
# alpha
t.out <- seq(from = 0, to = horizon, by = delta_t)
param.alpha <- ts(replicate(nb.sims, 0.5*(sigma^2)*(1 - exp(-a*t.out))^2/(a^2)), 
                  start = start(x), deltat = deltat(x))
alpha <- fwdrates + param.alpha
# The short-rate
r <- x + alpha
summary(as.vector(r))
summary(as.vector(x))
summary(as.vector(alpha))

```

```{r,eval=TRUE}
#x
ESGtoolkit::esgplotbands(x, xlab = "Times(years)", ylab = "x  quantiles", 
                         main = "x quantiles")
rg <- data.frame(x)
rg <- data.frame("x" = c(rg[,1]),"Sim" = 1,"Times" = times)
for (i in 2:nb.sims) {
  rg <- rbind(rg,data.frame("x" = c(x[,i]),
                            "Sim" = i,
                            "Times" = times))
}
rg %>%
  ggplot(aes(Times, x, group = factor(Sim))) +
  geom_line(aes(colour = factor(Sim)))

#alpha
rg <- data.frame(alpha)
rg <- data.frame("alpha" = c(rg[,1]),"Sim" = 1,"Times" = times)
rg %>%
  ggplot(aes(Times, alpha)) +
  geom_line()

ESGtoolkit::esgplotbands(r, xlab = "Times(years)", ylab = "short-rate quantiles", 
                         main = "short-rate quantiles")
# r
rg <- data.frame(r)
rg <- data.frame("r" = c(rg[,1]),"Sim" = 1,"Times" = times)
for (i in 2:nb.sims) {
  rg <- rbind(rg,data.frame("r" = c(r[,i]),
                            "Sim" = i,
                            "Times" = times))
}
rg %>%
  ggplot(aes(Times, r, group = factor(Sim))) +
  geom_line(aes(colour = factor(Sim)))
```
  
    
Now is time to compute the $R(t,T)$...
  
  
```{r,eval=TRUE}
B <- function(t,TT,a){
  rt <- -(1.0 - exp(-a*(TT - t)))/a
  return(rt)
}

phi <- function(t,a,sigma){
  phit <- sigma*sigma/(2.*a)*(1.- exp(-2.*a*t))
  return(phit)
}

system("rm HW.new.csv",ignore.stderr = T,wait = T)
TT <- times
Tmat <- seq(1/12,mat.max,1/12)
for (iSim in 1:nb.sims) {
  for (iT in 1:length(TT)) {
    rnew <- r[iT,iSim]
    #if(rnew<0) rnew=-rnew
    pp0 <- data.frame("starting.t.from.today.Years"=1,"Sim"=200,"Maturity.Years"=2,"Rate"=.0099)
    for (iT2 in 1:length(Tmat)){
      rates <- 1./(Tmat[iT2])*
        (
          B(TT[iT],TT[iT]+Tmat[iT2],a)*
            (-rnew+fwr[iT]+
             B(TT[iT],TT[iT]+Tmat[iT2],a)*phi(TT[iT],a,sigma)/2)+
          (TT[iT]+Tmat[iT2])*spr_asv(beta,(TT[iT]+Tmat[iT2]))-
          TT[iT]*spr_asv(beta,(TT[iT]+1.e-10))
      )
      
#Code if you are using Rquanlib estimation of forward rates      
#       rates <- 1./(Tmat[iT2])*(
#         B(TT[iT],TT[iT]+Tmat[iT2],a)*(rnew+curves$forwards[iT]+sigma^2/4*B(TT[iT],TT[iT]+Tmat[iT2],a)*B(0,(2*TT[iT]),a))-
#         (TT[iT]+Tmat[iT2])*curves$zerorates[(6*iT2)]+
#         TT[iT]*curves$zerorates[iT] 
#       )
      pp <- data.frame("starting.t.from.today.Years"=TT[iT],"Sim"=iSim,"Maturity.Years"=Tmat[iT2],"Rate"=rates)
      pp0=rbind(pp0,pp)
    }
    write.table(pp0[-1,],"HW.new.csv",row.names =F,append = T,col.names = F,dec=".",sep=",")
  }
}
```
  
Now we import saved simulations:

```{r,eval=TRUE}
library(dplyr)
library(ggplot2)
HW.new <- as.tbl(data.frame(read.table("HW.new.csv",sep=",",header = F,stringsAsFactors = F,dec = ".",colClasses = "numeric")))
names(HW.new)<-c("Years.from.Today","Simulation","Maturities.Years","Rates")
HW.new %>% head()
```
```{r,eval=FALSE,results=FALSE,echo=FALSE}
library(tidyr)
library(dplyr)
system("rm sim_*",ignore.stderr = T,wait = T)
HW.for.max <- HW.new%>%mutate(Rates.ok=Rates*100)%>%select(-Rates)%>%spread(Maturities.Years,Rates.ok)%>%arrange(Simulation)
for (i in 1:nb.sims){
 pp1<- HW.for.max %>%filter(Simulation==i)%>%filter(Years.from.Today>0)%>%select(-Simulation,-Years.from.Today)%>%write.table(paste("iswapall_",i-1,".dat",sep=""),sep=" ",row.names = F,col.names = F)
}
```

Now we show the results with a particular graph:

  - We do the mean of $R(t,T)$ over the `nb.sims`
  - We graph $R(t,T)$ over $T$ using $t$ as a label for different curves
  
```{r} 
HW.ok <- HW.new %>% 
  group_by(Years.from.Today,Maturities.Years) %>%
  summarise(Rate_mean = mean(Rates)) %>%
  mutate(Month.from.Today = Years.from.Today*12)


HW.ok %>%
  ggplot(aes(x = Maturities.Years,
             y = Rate_mean,
             colour = factor(Month.from.Today))) +
  geom_line()

```

  

